### 2015-05-29 chapters/terminology.tex
initial commit paper

The Unicode Cookbook for Linguists Managing orthographies in multilingual datasets using orthography profiles

Steven Moran \& Michael Cysouw

\textsc{Abstract}

In this paper, we define the notion of an orthography profile, which is a language-specific (or even resource-specific) description of the units and rules that are needed to adequately model a writing system. An orthography profile describes the Unicode code points, characters, graphemes and orthographic rules as used in a writing system. Our orthography profiles are used to describe writing systems and possibly even to transliterate them into another (for example IPA-like) transcriptions with the goal to harmonize orthographies across languages and resources on an ad-hoc basis. The central aspect of our approach is that we introduce an intermediate step of graphemic tokenization, which is closer to linguistic practice and easier to manage than transducer-based transliteration. We leverage orthography profiles to enable comparative analysis of languages with different writing systems and to perform quality control on datasets with highly specific orthographic systems. We also describe how our proposals relate to the current Unicode specification, and introduce some of the more delicate aspects of the Unicode Standard for a linguistic audience.

\textsc{Contents}

{[}{[}TOC{]}{]}

\section{1. Introduction}\label{introduction}

Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures. They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms. In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools). However, there is also a large community of users who take as many liberties in their writing as they can get away with.

For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out. As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44). However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language. More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation. By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group. And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school. The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).

The sometimes cumbersome and sometimes illogical structure, and the enormous variability of existing writing systems is a fact of life scholars have to accept and should try to adapt to as good as possible. Our plea here is a proposal for a formalization to do exactly that.

When considering the worldwide linguistic diversity, including all lesser-studied and endangered languages, there exist numerous different orthographies using symbols from the same scripts. For example, there are hundreds of orthographies using Latin-based alphabetic scripts. All of these orthographies use the same symbols, but these symbols differ in meaning and usage throughout the various orthographies. To be able to computationally use and compare different orthographies, we need a way to specify all orthographic idiosyncrasies in a computer-readable format (a process called `tailoring' in Unicode parlance). We call such specifications \textsc{orthography profiles}. Ideally, these specifications have to be integrated into so-called Unicode locale descriptions, though we will argue that in practice this is often not the most useful solution for the kind of problems arising in the daily practice of linguistics. Consequently, a central goal of this paper is to flesh out the linguistic challenges for locale descriptions, and work out suggestions to improve their structure for usage in a linguistic context. Conversely, we also aim to improve linguists' understanding and appreciation for the accomplishments of the Unicode Consortium in the development of the Unicode Standard.

The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison. Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al. 2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013). In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies. In many cases even different resources on the `same' language use different orthographic conventions. Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded. Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.

A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document. Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question. For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words. Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).

Although grapheme tokenization is only one part of the solution, it is an important and highly fruitful processing step. Given a grapheme tokenization, various subsequent tasks become easier, like (a) temporarily reducing the orthography in a processing pipeline, e.g.~only distinguishing high versus low vowels; (b) normalizing orthographies across sources (often including temporary reduction of oppositions), e.g.~specifying an (approximate) mapping to the International Phonetic Alphabet; (c) using co-occurrence statistics across different languages (or different sources in the same language) to estimate the probability of grapheme matches, e.g.~with the goal to find regular sound changes between related languages or transliterations between different sources in the same language.

Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3). We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard. A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4). As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).

The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists. First, we discuss the challenges of characterizing a writing system. To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6). Finally, we discuss practical issues with actual examples (Section 7). We provide reference implementation of our proposals in R and in Python, available as open-source libraries.

The following conventions are followed in this paper. All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association, 2005). Standard conventions are used for distinguishing between graphemic <\textgreater{}, phonemic / / and phonetic {[} {]} representations. For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014). Character names are represented in small capital letters (e.g.~LATIN SMALL LETTER SCHWA) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.

\section{2. Encoding}\label{encoding}

There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here. However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here. This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources, specifically http://www.unicode.org/history/ and various Wikipedia pages for the information presented here (when not references by traditional literature in the bibliography). A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003). Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.)

Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space. Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} ``distant writing'') have a long and widespread history since ancient times. The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al. 2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy. Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances. The `innovation' of those systems was that all characters of the written language were replaced one-to-one by visual signals. Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.) This of course implies that the usefulness of any such telegraphic encoding completely depends on the often rather haphazard structure of orthographic systems.

In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire. Originally, `bisignal' codes were used, consisting of two different signals. For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008: 282). More famous and influential, Samuel Morse in 1836 used long and short pulses. In those bisignal systems each character from the written language was encoded with a different number of signals (between one and five), so two different separators are needed: one between signals and one between characters (in Morse-code there is a short pause between signals and a long pause between characters).\footnote{Actually, Morse-code also includes an extra long pause between words. Interestingly, it took a long time to consider the written white-space as a bona-fide character that should simply be encoded with its own code point instead of with a different kind of pause. This happened only with the revision of the Baudot-code (see below) by Donald Murray in 1901, in which he finally introduced a white-space code. This principle has been followed ever since, and for example was included in the second version of the International Telegraph Alphabet (ITA2) in 1930.)

From those encodings, true \textsc{binary codes} developed with a fixed length of signals per character. In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems. An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as `self synchronizing'.) In the context of electric telegraphy, such a system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605.) There are 25 = 32 possible combination when using five binary signals; an encoding today designated as `5-bit'. These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits. As a solution, the Baudot code uses a so-called `shift' character, which signifies that from that point onwards (until shifted back) a different encoding is used, allowing for yet another set of 32 codes. In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet (ITA)} derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with 26 = 64 different possible characters). For decades, this encoding was the standard for all telegraphy and it is still in limited use today.

To also allow for different uppercase and lowercase letters and a large variety of control characters, the American Standards Association decided to propose a new 7-bit encoding in 1963 (with 27 = 128 different possible characters), known as the \textsc{American Standard Code for Information Interchange(ASCII)}, geared towards the encoding of English orthography. With the ascent of other orthographies in computer usage, the wish to encode further variation of Latin letters (like German <ß> or various letters with diacritics like <è>) led the Digital Equipment Corporation to introduce an 8-bit \textsc{Multinational Character Set (MCS)} (with 28 = 256 different possible characters), first used with the introduction of the VT220 Terminal in 1983. Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC 8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages. For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because 16 = 24, this means that ISO/IEC 8859 was actually a 8+4 = 12-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page). To be precise, ISO/IEC 8859 used the 7-bit ASCII as a basis for each code page, and defined 16 different 7-bit extensions, leading to (1+16) times 27 = 2,176 possible characters.)

This system almost immediately was understood to be insufficient and impractical, so various initiatives to extend and reorganize the encoding started in the 1980s. This led, for example, to various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple (e.g.~Mac OS Roman), which one still sometimes encounters today. More wide-ranging, various people in the 1980s started to develop true international code sets. In the United States, a group of computer scientists formed the Unicode Consortium, proposing a 16-bit encoding in 1991 (with 216 = 65,536 different possible characters). At the same time in Europe, the International Organization for Standardization (ISO) was working on ISO 10646 to supplant the ISO 8859 standard. Their first draft of the \textsc{Universal Character Set (UCS)} in 1990 was 31-bit (with theoretically 231 = 2,147,483,648 possible characters, but because of some technical restrictions only 679,477,248 were allowed). Since 1991, the Unicode Consortium and the International Organization for Standardization have jointly developed the \textsc{Unicode Standard}, or ISO/IEC 10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{Basic Multilingual Plane}, and 16 additional planes of 16-bit for further extensions (with in total 1+16 times 216 = 1,114,112 possible characters). The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters (Unicode Consortium 2014).

\section{3. Terminology}\label{terminology}

\subsection{The linguistic tradition}\label{the-linguistic-tradition}

In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard. Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way. The term `writing system' has two mutually exclusive meanings. First, it may refer to the way a particular language is written. In this sense the term refers to the writing system of a particular language, as, for example, in \emph{``the Serbian writing system uses two scripts: Latin and Cyrillic''}. Second, the term `writing system' may refer to a type of symbolic system as used among the world's languages to represent languages, as, for example, in \emph{``alphabetic writing system''}. In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{``the Latin and Cyrillic scripts are both alphabetic writing systems''}. To avoid confusion, this second notion of `writing system' would have been more aptly called \textsc{script system}. The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies. For example, both Serbian and Russian are written with subsets of the Cyrillic script. A single language, like Serbian or Japanese, can also be written using orthographies based on different scripts. Over the years linguists have typologized script systems in a variety of ways, with the tripartite classification of logographic, syllabic, and alphabetic remaining the most popular, even though there are at least half a dozen different types of script systems that can be distinguished (Daniels, 1990, 1996).

Only focussing here on the first sense, we distinguish two different kinds of writing systems as used to write a particular language, namely transcriptions and orthographies. First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level. It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined. A \textsc{transcription system} is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety. In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages. The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level. However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems. ) Note that IPA symbols are also often used to represent language on a phonemic level. However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation). Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes. Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm) and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP\footnote{http://asjp.clld.org/)).

Second, an \textsc{orthography} specifies the symbols, punctuations, and the rules in which a specific language is written in a standardized way. Orthographies are often based on a phonemic analysis, but they almost always include idiosyncrasies because of historical developments (like sound changes or loans) and because of the widely-followed principle of lexical integrity (i.e.~the attempt to write the same lexical root in a consistent way, also when synchronic phonemic rules change the pronunciation, for example with final devoicing in many Germanic languages). All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages. A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers. The mapping relation between phonemes and graphemes in practical orthographies is purposely shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme to a grapheme. Practical orthographies are intended to jumpstart written materials development by correlating a writing system with the sound units of a language (cf.~Meinhof \& Jones 1928). Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above). Further, a \textsc{transliteration} is a mapping between two different orthographies. It is the process of ``recording the graphic symbols of one writing system in terms of the corresponding graphic symbols of a second writing system'' (Kemp 2006:396). In straightforward cases, such a transliteration is simply a matter of replacing one symbol with another. However, there are widespread complications, like one-to-many or many-to-many mappings, which are not always easy, or even possible, to solve without listing all cases individually (cf.~Moran 2012, chp 2).

Breaking it down further, a script consists of \textsc{graphemes}, and graphemes consist of \textsc{characters}. In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term `character', i.e.~a conventional term for a unit in the Chinese writing system (Daniels 1996). This interpretation will not be further explored in this paper.) Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial. The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser. In practice, characters often consist of multiple building blocks, each of which could be considered a character in its own right. For example, although a Chinese character may be considered to be a single basic unanalyzable unit, at a more fine-grained level of analysis the internal structure of Chinese characters is often comprised of smaller semantic and phonetic units that should be considered characters (Sproat 2000). In alphabetic scripts, this problem is most forcefully exemplified by diacritics. A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters (Gaultney 2002). Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation (Daniels \& Bright 1996, xli). The central question is whether, for example, , , and should be considered four characters, or different combinations of three characters. In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.

For that reason, it is better to analyse an orthography as a collection of graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a particular writing system, alike to the phoneme is an abstract representation of a distinct sound in a specific language. The term `grapheme' was modeled after the term `phoneme' and represents a contrastive graphical unit in a writing system (see Kohrt 1986 for a historical overview of the term grapheme). Most importantly, a single grapheme regularly consists of multiple characters, like <th>, <ou> and <gh> in English (note that each character in these graphemes is also a separate grapheme in English). Such complex graphemes are often used to represent single phonemes. So, a combination of characters is used to represent a single phoneme. Note that the opposite is also found in writing systems, in cases in which a single character represents a combination of two or more phonemes. For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/. Further, conditioned or free variants of a grapheme are called \textsc{allographs}. For example, the distinctive forms of Greek sigma are conditioned, with <σ> being used word-internally and <ς> being used at the end of a word. In sum, there are many-to-many relationships between phonemes and graphemes as they are expressed in the myriad of language- and resource-specific orthographies.

@@

\subsection{The Unicode approach}\label{the-unicode-approach}

The conceptualisation and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium. For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange* *(ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded. This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font (Bird \& Simons 2003). In this situation, it was necessary to specify the font together with each document to ensure the rightful display of its content. To alleviate this problem of assigning different symbols to the same code points, in the late 80's and early 90's the Unicode Consortium set itself the ambitious goal of developing a single universal character encoding to provide a unique number, a code point, for every character in the world's writing systems. Nowadays, the Unicode Standard is the default encoding of the technologies that support the World Wide Web and for all modern operating systems, software and programming languages.

@@

Each version of the Unicode Standard (as of writing at version 7) consists of a set of specifications and guidelines that include (i) a core specification, (ii) code charts, (iii) standard annexes and (iv) a character database.\footnote{All documents of the Unicode Standard are available at http://www.unicode.org/versions/latest/. For a quick survey of the use of terminology inside the Unicode Standard, their glossary is particularly useful, available at http://www.unicode.org/glossary/. For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called ``General Structure'', is particularly insightful. Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world. ) The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text. The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)} are a set of technical standards that describe the implementation of the Unicode Standard for software development, Web standards, programming languages, etc. The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text files that describe the character properties, including a set of rich character and writing system semantics, for each character in the Unicode Standard. In this section, we introduce the basic Unicode concepts, but we will leave out many details. Please consult the above mentioned full documentation for a more detailed discussion. Further note that the Unicode Standard is exactly that, namely a standard. It normatively describes notions and rules to be followed. In the actual practice of applying this standard in a computational setting, a specific implementation is necessary. The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org )

The Unicode Standard is a \textsc{character encoding system} which goal it is to support the interchange and processing of written characters and text in a computational setting. Underlyingly, the character encoding is represented by a range of numerical values called a \textsc{code space}, which is used to encode a set of characters. A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range). In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering 0·163 + 0·162 + 7·161 + 0·160 = 7·16 = 112. Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, i.e.~112 = 1·27 + 1·26 + 1·25 + 0·24 + 0·23 + 0·22 + 0·21 + 0·20 = 64 + 32 + 16.) That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and the context in which that character appears.

>>>>

\section{The linguistic terminological tradition}
\label{the-linguistic-tradition}

In this section we provide a very brief overview of the linguistic terminology concerning writing systems before turning to the slightly different computational terminology in the next section on the Unicode Standard.

Linguistically speaking, a \textsc{writing system} is a symbolic system that uses visible or tactile signs to represent language in a systematic way. The term `writing system' has two mutually exclusive meanings. First, it may refer to the way a particular language is written. In this sense the term refers to the writing system of a particular language, as, for example, in \emph{`the Serbian writing system uses two scripts: Latin and Cyrillic'}. Second, the term `writing system' may also refer to a type of symbolic system as used among the world's languages to represent the language, as, for example, in \emph{`alphabetic writing system'}. In this latter sense the term refers to how scripts have been classified according to the way that they encode language, as in, for example, \emph{`the Latin and Cyrillic scripts are both alphabetic writing systems'}. To avoid confusion, this second notion of `writing system' would more aptly have been called `script system'.

The term \textsc{script} refers to a collection of distinct symbols as employed by one or more orthographies. For example, both Serbian and Russian are written with subsets of the Cyrillic script. A single language, like Serbian or Japanese, can also be written using orthographies based on different scripts. Over the years linguists have typologized script systems in a variety of ways, with the tripartite classification of logographic, syllabic, and alphabetic remaining the most popular, even though there are at least half a dozen different types of script systems that can be distinguished (Daniels, 1990, 1996).

Only focussing here on the first sense of writing systems, we distinguish two different kinds of writing systems used to write a particular language, namely transcriptions and orthographies. First, \textsc{transcription} is a scientific procedure (and also the result of that procedure) for graphically representing the sounds of human speech at the phonetic level. It incorporates a set of unambiguous symbols to represent speech sounds, including conventions that specify how these symbols should be combined. A transcription system is a specific system of symbols and rules used for transcription of the sounds of a spoken language variety. In principle, a transcription system should be language-independent, in that it should be applicable to all spoken human languages. The International Phonetic Alphabet (IPA) is a commonly used transcription system that provides a medium for transcribing languages at the phonetic level. However, there is a long history of alternative kinds of transcription systems (see Kemp 1994) and today various alternatives are in widespread use (e.g.~X-SAMPA and Cyrillic-based phonetic transcription systems).\footnote{Most users of IPA do not follow the standard explicitly and many dialects based on it have emerged, e.g.~the Africanist and Americanist transcription systems.} Note that IPA symbols are also often used to represent language on a phonemic level. However, it is important to realize that in this usage the IPA symbols are not a transcription system, but rather an orthography (though with strong links to the pronunciation). Further, a transcription system does not need to be highly detailed, but it can also be a system of broad sound classes. Although such an approximative transcription is not normally used in linguistics, it is widespread in technological approaches (e.g.~Soundex, Beider \& Morse 2010)\footnote{http://stevemorse.org/phonetics/bmpm.htm , http://stevemorse.org/phonetics/bmpm2.htm} and it is sometimes fruitfully used in automatic approaches to historical linguistics (Dolgopolsky 1986, ASJP).\footnote{http://asjp.clld.org/}

Second, an \textsc{orthography} specifies the symbols, punctuations, and the rules in which a specific language is written in a standardized way. Orthographies are often based on a phonemic analysis, but they almost always include idiosyncrasies because of historical developments (like sound changes or loans) and because of the widely-followed principle of lexical integrity (i.e.~the attempt to write the same lexical root in a consistent way, also when synchronic phonemic rules change the pronunciation, for example with final devoicing in many Germanic languages). All orthographies are language-specific (and often even resource-specific), although individual symbols or rules might be shared between languages.

A \textsc{practical orthography} is a strongly phoneme-based writing system designed for practical use by speakers. The mapping relation between phonemes and graphemes in practical orthographies is purposely shallow, i.e.~there is mostly a systematic and faithful mapping from a phoneme to a grapheme. Practical orthographies are intended to jumpstart written materials development by correlating a writing system with the sound units of a language (cf.~Meinhof \& Jones 1928). Symbols from the IPA are often used by linguists in the development of practical orthographies for languages without writing systems, though this usage of IPA symbols should not be confused with transcription (as defined above).

Further, a \textsc{transliteration} is a mapping between two different orthographies. It is the process of ``recording the graphic symbols of one writing system in terms of the corresponding graphic symbols of a second writing system'' (Kemp 2006:396). In straightforward cases, such a transliteration is simply a matter of replacing one symbol with another. However, there are widespread complications, like one-to-many or many-to-many mappings, which are not always easy, or even possible, to solve without listing all cases individually (cf.~Moran 2012, chp 2).

Breaking it down further, a script consists of \textsc{graphemes}, and graphemes consist of \textsc{characters}. In the linguistic terminology of writing systems, a \textsc{character} is a general term for any self-contained element in a writing system.\footnote{There is a second interpretation of the term `character', i.e.~a conventional term for a unit in the Chinese writing system (Daniels 1996). This interpretation will not be further explored in this paper.} Although in literate societies most people have a strong intuition about what the characters are in their particular orthography or orthographies, it turns out that the separation of an orthography into separate characters is far from trivial. The widespread intuitive notion of a character is strongly biased towards educational traditions, like the alphabet taught at schools, and technological possibilities, like the available type pieces in a printer's job case, the keys on a typewriter, or the symbols displayed in Microsoft Word's symbol browser. In practice, characters often consist of multiple building blocks, each of which could be considered a character in its own right. For example, although a Chinese character may be considered to be a single basic unanalyzable unit, at a more fine-grained level of analysis the internal structure of Chinese characters is often comprised of smaller semantic and phonetic units that should be considered characters (Sproat 2000). In alphabetic scripts, this problem is most forcefully exemplified by diacritics.

A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or through other characters (Gaultney 2002). Diacritics are sometimes used to distinguish homophonous words, but they are more often used to indicate a modified pronunciation (Daniels \& Bright 1996, xli). The central question is whether, for example, <e>, <è>, <a> and <à> should be considered four characters, or different combinations of three characters. In general, multiple characters together can form another character, and it is not always possible to decide on principled grounds what should be the basic building blocks of an orthography.

For that reason, it is better to analyse an orthography as a collection of graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a particular writing system, alike to the phoneme is an abstract representation of a distinct sound in a specific language. The term `grapheme' was modeled after the term `phoneme' and represents a contrastive graphical unit in a writing system (see Kohrt 1986 for a historical overview of the term grapheme). Most importantly, a single grapheme regularly consists of multiple characters, like <th>, <ou> and <gh> in English (note that each character in these graphemes is also a separate grapheme in English). Such complex graphemes are often used to represent single phonemes. So, a combination of characters is used to represent a single phoneme. Note that the opposite is also found in writing systems, in cases in which a single character represents a combination of two or more phonemes. For example, <x> in English orthography represents a combination of the phonemes /k/ and /s/.

Further, conditioned or free variants of a grapheme are called \textsc{allographs}. For example, the distinctive forms of Greek sigma are conditioned, with <σ> being used word-internally and <ς> being used at the end of a word. In sum, there are many-to-many relationships between phonemes and graphemes as they are expressed in the myriad of language- and resource-specific orthographies.

@@

\section{The Unicode approach}
\label{the-unicode-approach}

The conceptualization and terminology of writing systems was rejuvenated through the development of the Unicode Standard, with major input from Mark Davis, co-founder and long-term president of the Unicode Consortium. For many years, `exotic' writing systems and phonetic transcription systems on personal computers were constrained by the American Standard Code for Information Interchange (ASCII) character encoding scheme (based on the Latin script), which only allowed for a strongly limited number of different symbols to be encoded. This implied that users could either use and adopt the (extended) Latin alphabet or they could assign new symbols to the small number of code points in the ASCII encoding scheme to be rendered by a specifically designed font (Bird \& Simons 2003). In this situation, it was necessary to specify the font together with each document to ensure the rightful display of its content. To alleviate this problem of assigning different symbols to the same code points, in the late 80's and early 90's the Unicode Consortium set itself the ambitious goal of developing a single universal character encoding to provide a unique number, a code point, for every character in the world's writing systems. Nowadays, the Unicode Standard is the default encoding of the technologies that support the World Wide Web and for all modern operating systems, software and programming languages.

@@

Each version of the Unicode Standard (as of writing at version 7) consists of a set of specifications and guidelines that include (i) a core specification, (ii) code charts, (iii) standard annexes and (iv) a character database.\footnote{All documents of the Unicode Standard are available at http://www.unicode.org/versions/latest/. For a quick survey of the use of terminology inside the Unicode Standard, their glossary is particularly useful, available at http://www.unicode.org/glossary/. For a general introduction to the principles of Unicode, Chapter 2 of the core specification, called ``General Structure'', is particularly insightful. Different from many other documents of the Unicode Standard, this general introduction is relatively easy to read and illustrated with many interesting examples from various orthography traditions all over the world.} The \textsc{core specification} is a book directed toward human readers that describes the formal standard for encoding multilingual text. The \textsc{code charts} provide a humanly readable online reference to the character contents of the Unicode Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)} are a set of technical standards that describe the implementation of the Unicode Standard for software development, Web standards, programming languages, etc. The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text files that describe the character properties, including a set of rich character and writing system semantics, for each character in the Unicode Standard. In this section, we introduce the basic Unicode concepts, but we will leave out many details. Please consult the above mentioned full documentation for a more detailed discussion. Further note that the Unicode Standard is exactly that, namely a standard. It normatively describes notions and rules to be followed. In the actual practice of applying this standard in a computational setting, a specific implementation is necessary. The most widely used implementation of the Unicode Standard is the \textsc{International Components for Unicode (ICU)}, which offers C/C++ and Java libraries trying to implement to Unicode Standard.\footnote{http://icu-project.org}

The Unicode Standard is a \textsc{character encoding system} which goal it is to support the interchange and processing of written characters and text in a computational setting. Underlyingly, the character encoding is represented by a range of numerical values called a \textsc{code space}, which is used to encode a set of characters. A \textsc{code point} is a unique non-negative integer within a code space (i.e.~within a certain numerical range). In the Unicode Standard character encoding system, an \textsc{abstract character}, for example a ``LATIN SMALL LETTER P'', is mapped to a particular code point, such as the decimal value 112, normally represented in hexadecimal and in Unicode parlance as (U+)0070.\footnote{Hexadecimal (`base-16') 0070 is equivalent to decimal (`base-10') 112, which can be calculated by considering that $(0\cdot16^3) + (0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly, computers will of course treat this code point binary (`base-2') as 11100000, as can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) + (0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 + 16 = 112$.} That encoded abstract character is rendered on a computer screen (or printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and the context in which that character appears.
